---
title: 'The Spatial Distribution of Farmers Markets in Philadelphia'
author: "Zhanchao Yang, Haoyu Zhu, Kavana Raju"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide
    code_download: yes
    mathjax: default
---
```{r load packages, message=FALSE, warning=FALSE, include=FALSE}

options(scipen=999)
options(digits = 3)

library(tidyverse)
library(sf)
library(here)
library(kableExtra)
library(patchwork)
library(spatstat.explore)
library(spatstat)
library(proxy)
```

```{r, include=FALSE}
market <- read_sf("data/Philadelphia_Farmers_Markets201302.shp")
zipcode <- read_sf("data/Philadelphia_ZipCodes.shp")
philly <- read_sf("data/Philadelphia.shp")
```


# Introduction

Access to fresh and healthy food is a critical issues in many American cities. Lack of access to fresh food may lead to significant public health challenges and worsen health disparities. In Philadelphia, the Food Trust has established a network of farmers markets to improve access to fresh food. However, some neighborhoods may still face limited or no access to these markets, raising significant concerns about the potential food desert and access inequity. This study aims to analyze the spatial distribution of farmers markets in Philadelphia to determine whether they are randomly placed, clustered, or dispersed across the city. At the same time, the study also try to identify the areas lack of access to farmers markets and the potential food deserts.

# Methodology

# Results

```{r, message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = philly, fill = "grey80") +
  geom_sf(data = zipcode, fill = NA, color = "white") +
  geom_sf(data = market, color = "#c44536", size = 1.5) +
  theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.4)
        ) +
  labs(title = "Farmers Markets in Philadelphia")
```
```{r, warning=FALSE, message=FALSE}
philly_window <- as.owin(st_transform(philly, crs = st_crs(market)))
market_coords <- st_coordinates(market)
market_pp <- ppp(x = market_coords[,1], y = market_coords[,2], window = philly_window)
```

```{r}
# Nearest neighbor analysis
nnd <- nndist.ppp(market_pp)

# Average Observed Distance
MeanObsDist <- mean(nnd)

# Average Expected Distance
# The expected mean nearest neighbor distance under Complete Spatial Randomness (CSR).
n <- npoints(market_pp)
area <- area.owin(market_pp$window)
MeanExpDist <- 0.5 / sqrt(n / area)

#Standard Error
SE <- 0.26136 / sqrt(n*n / area)
```

```{r}
NNI <- MeanObsDist / MeanExpDist # Nearest Neighbor Index
zscore <- (MeanObsDist - MeanExpDist)/SE  #Calculating the z-score
pval<-ifelse(zscore > 0, 1 - pnorm(zscore), pnorm(zscore))  #Calculating the p-value

results <- data.frame(
  Metric = c("Nearest Neighbor Index (NNI)", "Z-Score", "P-Value"),
  Value = c(round(NNI, 3), round(zscore, 3), round(pval, 7))
)

results %>%
  kable("html", col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
```{r}
max.distance <- max(proxy::dist(
  data.frame(cbind(x = market_coords[,1], y = market_coords[,2])),
  data.frame(cbind(x = market_coords[,1], y = market_coords[,2])),
  method = "euclidean"
))

cat("Maximum Distance:", max.distance, "\n")
```

```{r, message=FALSE, warning=FALSE}
khat <-Kest(market_pp, rmax=28000, correction="Ripley")
khat_df <- data.frame(
  r = khat$r,             # Distance values
  iso = khat$iso,         # Observed K(r) (isotropic)
  theo = khat$theo        # Theoretical K(r)
)

ggplot(khat_df, aes(x = r)) +
  geom_line(aes(y = iso, color = "Observed K(r)"), size = 2, color = "#197278") +
  geom_line(aes(y = theo, color = "Theoretical K(r)"), linetype = "dashed", size = 1, color = "#c44536") +
  labs(
    x = "r (Distance)",
    y = "Ripley's K-Function",
    title = "Ripley's Estimated K-Function",
    color = "Legend"
  ) +
theme_light() +
theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        axis.text.x=element_text(size=6),
        axis.text.y=element_text(size=6),
        axis.title=element_text(size=8))
```
```{r}
khat_df$difference <- khat_df$iso - khat_df$theo
threshold_index <- which(khat_df$difference > 0)[1]

if (!is.na(threshold_index)) {
  consistent_start <- khat_df$r[threshold_index]
  message("The observed K(r) is consistently higher than the theoretical K(r) starting at r = ", consistent_start)
} else {
  message("The observed K(r) does not consistently exceed the theoretical K(r) within the given range.")
}
```

```{r}
Kenv <- spatstat.explore::envelope(market_pp,fun="Kest", rmax=28000, nsim=9, nrank=1)
```
```{r}
Kenv_df <- data.frame(
  r = Kenv$r,              # Distance values
  obs = Kenv$obs,          # Observed K-function
  lower = Kenv$lo,         # Lower envelope
  upper = Kenv$hi          # Upper envelope
)

ggplot(Kenv_df, aes(x = r)) +
  geom_line(aes(y = obs, color = "Observed K(r)"), size = 2, color = "#772e25") +
  geom_line(aes(y = lower, color = "Lower Envelope"), linetype = "dashed", size = 1, color = "#197278") +
  geom_line(aes(y = upper, color = "Upper Envelope"), linetype = "dashed", size = 1, color = "#c44536") +
  labs(
    x = "r (Distance)",
    y = "Khat(r)",
    title = "Ripley's Khat with Confidence Envelopes",
    color = "Legend"
  ) +
  theme_light() +
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        axis.text.x=element_text(size=6),
        axis.text.y=element_text(size=6),
        axis.title=element_text(size=8))
```
```{r}
Kenv_df$difference_lower <- Kenv_df$obs - Kenv_df$lower
first_below_index <- which(Kenv_df$difference_lower < 0)[1]

if (!is.na(first_below_index)) {
  below_start <- Kenv_df$r[first_below_index]
  message("The observed K(r) falls below the lower envelope starting at r = ", below_start)
} else {
  message("The observed K(r) does not fall below the lower envelope within the given range.")
}
```

```{r}
lhat <- Lest(market_pp, rmax=28000, correction="Ripley")

lhat_df <- data.frame(
  r = lhat$r,                      # Distance values
  L_obs = lhat$iso - lhat$r,       # Observed L-function minus r
  L_theo = lhat$theo - lhat$r      # Theoretical L-function minus r
)

ggplot(lhat_df, aes(x = r)) +
  geom_line(aes(y = L_obs, color = "Observed L(r)"), size = 2, color = "#197278") +
  geom_line(aes(y = L_theo, color = "Theoretical L(r)"), linetype = "dashed", size = 1, color = "#c44536") +
  labs(
    x = "r (Distance)",
    y = "Ripley's L - r",
    title = "Ripley's Estimated L-Function",
    color = "Legend"
  ) +
  theme_light() +
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        axis.text.x=element_text(size=6),
        axis.text.y=element_text(size=6),
        axis.title=element_text(size=8))
```

```{r}
Lenv <- spatstat.explore::envelope(market_pp,fun="Lest", rmax=28000, nsim=9,nrank=1)
```

```{r}
L2 <- Lenv
L2_df <- data.frame(
  r = L2$r,                   # Distance values
  obs = L2$obs - L2$r,        # Adjusted observed L-function
  theo = L2$theo - L2$r,      # Adjusted theoretical L-function
  lo = L2$lo - L2$r,          # Lower confidence envelope
  hi = L2$hi - L2$r           # Upper confidence envelope
)

ggplot(L2_df, aes(x = r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80", alpha = 0.5) +
  geom_line(aes(y = obs, color = "Observed L(r)"), size = 2, color = "#197278") +
  geom_line(aes(y = theo, color = "Theoretical L(r)"), linetype = "dashed", size = 1, color = "#c44536") +
  labs(
    x = "r (Distance)",
    y = "L(r) - r",
    title = "Ripley's L-Function with Confidence Envelopes",
    color = "Legend"
  ) +
  theme_light() +
  theme(plot.subtitle = element_text(size = 9,face = "italic"),
        plot.title = element_text(size = 12, face = "bold"),
        axis.text.x=element_text(size=6),
        axis.text.y=element_text(size=6),
        axis.title=element_text(size=8))
```
```{r}
ggplot() +
  geom_sf(data = zipcode, aes(fill = Pop2000), color = "white") +
   scale_fill_continuous(low = "#FAF9F6", high = "#c44536", name= "Population") +
  theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(size = 12, face = "bold"),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill=NA, size=0.4)
        ) +
  labs(title = "Philadelphia Population by Zip Code")
```
# Discussion

```{r}
ggplot() +
  geom_sf(data = zipcode, aes(fill = MedIncome), color = "white") +
  scale_fill_continuous(low = "#FAF9F6", high = "#197278", name = "Median Income") +
    geom_sf(data = market, aes(), color = "#c44536", size = 2) +
    theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(size = 12, face = "bold"),
    panel.background = element_blank(),
    panel.border = element_rect(colour = "grey", fill = NA, size = 0.4)
  ) +
    labs(
    title = "Farmers Markets and Median Household Income by Zip Code")
```
